---
title: "README"
author: "Miles Nienstadt & Lily Nathan"
output: 
  github_document:
  pandoc_args: ["--wrap=none"]
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# install.packages(c("tidyverse", "knitr", "kableExtra"))  # run once, then leave commented out
library(tidyverse)
library(knitr)
library(kableExtra)
```

# Getting into Business: Real Estate Investment Data Exploration

**Dataset:** FHFA House Price Index (HPI)

## 1. Understanding the Data

### When was the data acquired?

The dataset covers 1975 to 2026 (all-transactions HPI data starts in 1975; the other four HPI types begin in 1991), with monthly and quarterly observations continuing through the most recent period available. FHFA updates this file on a rolling basis; for reproducibility, we note the date we pulled it (rather than a fixed "last updated" date, since the source updates monthly/quarterly).

### Where was the data acquired?

The data was obtained from the Federal Housing Finance Agency (FHFA), publicly available at <https://www.fhfa.gov/hpi/download/monthly/hpi_master.csv>. Geographically it covers the entire United States plus Puerto Rico, broken into four levels: USA/Census Division, State, MSA (metropolitan statistical area), and Puerto Rico — 472 unique geographic places in total.

### How was the data acquired?

FHFA constructs the HPI using a weighted, repeat-sales methodology: it tracks price changes by comparing repeat sales or refinancings of the same single-family properties over time, rather than comparing different homes to each other. The underlying transactions come from mortgages purchased or securitized by Fannie Mae and Freddie Mac, so the data reflects conforming-loan mortgage activity rather than all real estate sales (cash sales and jumbo loans are excluded).

### What are the attributes of this dataset, and what data types are they?

| Column | Type | Description |
|---|---|---|
| `hpi_type` | Nominal (categorical) | Which HPI variant: traditional, non-metro, manufactured, distress-free, developmental |
| `hpi_flavor` | Nominal (categorical) | Underlying data used: purchase-only, all-transactions, expanded-data |
| `frequency` | Nominal (categorical) | monthly or quarterly |
| `level` | Nominal (categorical) | Geography level: USA or Census Division, State, MSA, Puerto Rico |
| `place_name` | Nominal (categorical) | Name of the geography (e.g., "Abilene, TX") |
| `place_id` | Nominal (categorical) | Abbreviation or CBSA code identifying the place |
| `yr` | Interval (numeric) | Calendar year, 1975–2026 |
| `period` | Ordinal (numeric) | Month (1–12) or quarter (1–4), depending on frequency |
| `index_nsa` | Ratio (numeric) | Index value, not seasonally adjusted (base = 100) |
| `index_sa` | Ratio (numeric) | Index value, seasonally adjusted (base = 100); only computed for some series |
| `rstderr` | Ratio (numeric) | Standard error of the index estimate, where available |
| `note` | Nominal (text) | Footnote/annotation on the observation, where present |

#### Understanding the key variables in more depth

HPI stands for House Price Index. FHFA (the Federal Housing Finance Agency) publishes several versions of this index, which differ along two independent dimensions captured by `hpi_type` and `hpi_flavor` — type describes what the index measures or which properties it covers, while flavor describes which underlying transactions feed into it.

**`hpi_type` — what the index measures**

- **traditional (177,642 rows):** the standard HPI reported in FHFA press releases; a general measure of single-family price movement with no properties excluded.
- **non-metro (5,922 rows):** covers non-metropolitan (rural) areas within a state, i.e. areas outside any MSA.
- **distress-free (1,988 rows):** removes bank-owned (REO) sales and short sales from the purchase-only data before estimating the index, isolating price trends from distressed-sale discounts.
- **developmental (247 rows):** newer, experimental index types (e.g., for Puerto Rico) that FHFA considers less mature/reliable than the traditional index.
- **manufactured (212 rows):** built only from conventional mortgages on single-family detached manufactured homes; also considered developmental.

**`hpi_flavor` — which underlying transactions are used**

- **purchase-only (29,338 rows):** based only on actual home purchase transactions financed with conforming mortgages.
- **all-transactions (89,791 rows):** adds appraisal values from refinance mortgages to the purchase-only sample, giving broader coverage (this is why it's available for smaller/rural geographies where purchase counts alone would be too thin).
- **expanded-data (66,882 rows):** adds sales-price data from county recorder offices and FHA-backed mortgages to the purchase-only sample; this is the flavor FHFA uses to set annual conforming loan limits.

**`place_name` / `place_id` — what geographies are actually in the data**

The `level` column tells you the geographic scale, and `place_name`/`place_id` together identify the specific location at that scale. There is no ZIP-code-level data in this file — the finest geography available is the MSA (metro area).

- **USA or Census Division** (10 places): the national index plus the 9 Census Bureau regional divisions, e.g. "East North Central Division" (`place_id` DV_ENC), or "United States" (`place_id` USA).
- **State** (51 places): all 50 states plus Washington, D.C., identified by 2-letter postal abbreviation, e.g. "Alabama" (`place_id` AL).
- **MSA** (410 places): individual Metropolitan Statistical Areas, identified by their 5-digit CBSA code, e.g. "Abilene, TX" (`place_id` 10180).
- **Puerto Rico** (1 place): a single territory-wide developmental index.

```{r load-data}
# load data directly from FHFA for reproducibility
hpi_master <- read.csv("https://www.fhfa.gov/hpi/download/monthly/hpi_master.csv")

# structure
str(hpi_master)
```

```{r explore-categories}
# unique categories for the key categorical fields
sapply(hpi_master[c("hpi_type","hpi_flavor","frequency","level")], unique)

# year range and number of unique places
range(hpi_master$yr)
n_distinct(hpi_master$place_id)
```

```{r type-flavor-geo-breakdown}
# type/flavor/geography breakdowns
table(hpi_master$hpi_type)
table(hpi_master$hpi_flavor)
table(hpi_master$hpi_type, hpi_master$hpi_flavor)

hpi_master %>%
  distinct(level, place_name, place_id) %>%
  count(level)
```
### Missing values

```{r missing-values}
missing_summary <- hpi_master %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Column", values_to = "Missing_Count") %>%
  mutate(Missing_Pct = round(100 * Missing_Count / nrow(hpi_master), 1)) %>%
  arrange(desc(Missing_Count))

kable(missing_summary, caption = "Missing Values by Column")
```

Note: `index_sa` (seasonally adjusted) is missing for about 48% of rows (89,897 of 186,011) — this is because seasonal adjustment isn't computed for all `hpi_type`/`level` combinations, not random missingness. The same is true for `rstderr` and `note`, which are populated only for certain series/annotations.

### Handling Missing Values
Missing values in this dataset are largely structural rather than random. For index_sa, seasonal adjustment is not available for every series or geographic level. Because of this, our primary analysis will use index_nsa, which has much broader coverage. If seasonally adjusted values are needed, we can restrict the analysis to subsets of the data where index_sa is available. Another possible approach would be to estimate seasonal adjustments for missing observations using an appropriate time-series method, though this would require additional assumptions.

For rstderr, we would not attempt to manually fill in missing values. This variable reflects the estimated uncertainty or reliability of the HPI calculation, so imputing it could create misleading measures of precision. Instead, missing rstderr values should be left as missing and only used where FHFA reports them.

## 2. Data Summary & Initial Insights

### Summary statistics for numeric variables

```{r numeric-summary}
numeric_summary <- hpi_master %>%
  summarise(across(
    c(yr, period, index_nsa, index_sa, rstderr),
    list(
      Mean   = ~mean(.x, na.rm = TRUE),
      Median = ~median(.x, na.rm = TRUE),
      SD     = ~sd(.x, na.rm = TRUE),
      Min    = ~min(.x, na.rm = TRUE),
      Max    = ~max(.x, na.rm = TRUE),
      NAs    = ~sum(is.na(.x))
    ),
    .names = "{.col}__{.fn}"
  )) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "__") %>%
  pivot_wider(names_from = Stat, values_from = value)

kable(numeric_summary, digits = 2, caption = "Summary Statistics for Numeric Variables")
```

### Frequency tables for categorical variables

```{r frequency-tables}
kable(hpi_master %>% count(hpi_type, sort = TRUE), caption = "Observations by hpi_type")
kable(hpi_master %>% count(hpi_flavor, sort = TRUE), caption = "Observations by hpi_flavor")
kable(hpi_master %>% count(level, sort = TRUE), caption = "Observations by Geography Level")
kable(hpi_master %>% count(frequency, sort = TRUE), caption = "Observations by Frequency")
```

### Visualization: distribution of index_nsa

```{r hist-index}
ggplot(hpi_master, aes(x = index_nsa)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Non-Seasonally-Adjusted HPI",
       x = "Index (NSA)", y = "Count") +
  theme_minimal()
```

### Visualization: National HPI Trend Over Time

```{r box-index-by-type}
national_over_time <- hpi_master %>%
  filter(
    place_id == "USA",
    hpi_type == "traditional",
    hpi_flavor == "purchase-only",
    frequency == "monthly"
  ) %>%
  mutate(date = as.Date(paste(yr, period, "01", sep = "-")))

ggplot(national_over_time, aes(x = date, y = index_nsa)) +
  geom_line(color = "steelblue", linewidth = 0.9) +
  labs(
    title = "U.S. National House Price Index Over Time",
    subtitle = "Traditional, Purchase-Only, Monthly (Not Seasonally Adjusted)",
    x = NULL,
    y = "Index (NSA, base = 100)"
  ) +
  theme_minimal()
```

### Visualization: Census Division Trends

```{r census-trend}
division_trends <- hpi_master %>%
  filter(
    level == "USA or Census Division",
    place_id != "USA",
    hpi_type == "traditional",
    hpi_flavor == "purchase-only",
    frequency == "quarterly"
  ) %>%
  mutate(
    date = as.Date(paste(yr, period * 3, "01", sep = "-"))
  )

ggplot(
  division_trends,
  aes(x = date, y = index_nsa, color = place_name)
) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "House Price Trends Across U.S. Census Divisions",
    subtitle = "Traditional Purchase-Only HPI",
    x = NULL,
    y = "House Price Index",
    color = "Census Division"
  ) +
  theme_minimal()
```

### Five-Year House Price Growth by State

```{r year-change}
state_data <- hpi_master %>%
  filter(
    level == "State",
    hpi_type == "traditional",
    hpi_flavor == "purchase-only",
    frequency == "quarterly"
  )

latest_year <- max(state_data$yr, na.rm = TRUE)

latest_period <- state_data %>%
  filter(yr == latest_year) %>%
  summarise(period = max(period, na.rm = TRUE)) %>%
  pull(period)

state_latest <- state_data %>%
  filter(
    yr == latest_year,
    period == latest_period
  ) %>%
  select(place_name, current_hpi = index_nsa)

state_5yr <- state_data %>%
  filter(
    yr == latest_year - 5,
    period == latest_period
  ) %>%
  select(place_name, hpi_5yr = index_nsa)

state_growth <- state_latest %>%
  inner_join(state_5yr, by = "place_name") %>%
  mutate(
    pct_change = 100 * (current_hpi / hpi_5yr - 1)
  )

ggplot(
  state_growth,
  aes(x = reorder(place_name, pct_change), y = pct_change)
) +
  geom_col(fill = "steelblue", width = 0.8) +
  coord_flip() +
  labs(
    title = "Five-Year House Price Growth by State",
    subtitle = "Traditional Purchase-Only HPI",
    x = NULL,
    y = "5-Year Change (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 11)
  )
```

### 3. Complementary Dataset: Census Building Permits Survey

A useful complementary dataset is the U.S. Census Bureau Building
Permits Survey.

The Building Permits Survey would be useful because it provides
information on new privately owned residential construction across
different geographic areas. It includes data at the national, state,
metropolitan, county, and local levels, which makes it useful for
examining changes in housing supply over time.

This dataset complements the FHFA House Price Index because the HPI
shows how home prices change, while the Building Permits Survey provides
information about new housing supply. Combining the two datasets could
help us examine whether areas with more new construction experience
different house-price trends than areas where housing supply is growing
more slowly.

Since both datasets include state and metropolitan-area data, they could
also be compared geographically to better understand differences between
housing markets.

Source: [U.S. Census Bureau Building Permits
Survey](https://www.census.gov/construction/bps/index.html)






